---
title: "SMPA - Grouping Assignment"
author: "YourNameHere"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
---

```{r}

#load libraries
library(tidyverse)
library(lubridate)

#load data
payments <- readRDS("data/trump_campaign_payments.rds")

```

## Introduction

**What data will you be working with?**

A dataset of disbursements (payments and purchases) by the Donald J. Trump for President campaign committee during the 2020 election cycle. The included payments are from 2017 and 2018.

This data and examples were developed as part of the NYT Open training, which we've adapted from here. Note that the dataset has been alterted for instructional purposes, so doesn't represent the original FEC data in its entirety.

In the environment tab in the upper right, click on the `payments` dataset to open the RStudio viewer to examine the data. You can also run the chunk below as well to bring up a slice of the data.

```{r}

head(payments)


```

## Before you begin

Take a moment to look at the various categories included for each disbursement.

-   Notice that there are several date-related fields:

    -   FEC election cycle

    -   report year

    -   election type

    -   disbursement date

-   There are a few fields about the recipient. Along with city and state, you have:

    -   the recipient name

    -   the payee's first and last name (an individual who was paid for the "recipient") and

    -   the entity type (individual v. organization)

## Initial Narrative Questions

Without doing any analysis yet, what are three questions you might try to answer using this dataset? Write your narrative answers in the spaces below.

1.  What has been the largest and the least amount of disbursement during the campaign? 

2.  What was the description of the disbursement for which the campaign spent the most money?  

3.  What was the proportion of money spent in organizations versus spent in individuals? 

## Coding Assignment Questions

**Please write your R tidyverse code to help answer the questions below in the provided R chunks. You can make more than one chunk if you prefer.**

**Also, describe what you're doing either in text out "in the open" above the chunk, or in comments (prefaced by #) in the R chunk itself.**

**You'll need to use grouping, filtering, or a combination of both to find the answers. Remember that R is case-sensitive when you refer to any text values within the columns**

<br>

How much was spent in the 2017 vs. 2018 reporting year?

```{r}

#I filtered 2 sections of the payments dataset by year. First, 2017 and then 2018. 
#Then, I grouped by the commitee name and report year 
#Then, I used the sumarise formula to sum the disbursement amount for each year

payments %>% 
   filter(report_year == 2017) %>% 
   group_by(committee_name, report_year) %>% 
  summarise(sum(disbursement_amount, na.rm = TRUE))

#Amount spent in the 2017 reporting year was 16888149	

payments %>% 
   filter(report_year == 2018) %>% 
   group_by(committee_name, report_year) %>% 
  summarise(sum(disbursement_amount, na.rm = TRUE))

#Amount spent in the 2018 reporting year was 37406905	


```

Tally the total overall disbursement amounts for each month (distinguishing between say, Oct. 2018 v. Oct. 2019). In which month was spending the highest?

```{r}

#I grouped it by month and year of disbursement and summarised the respective amount for each combination
#of month and year. Then, I used the arrange for the descending order of the disbursement amount to 
#find the highest spending in the respective month & year combination

payments %>%
  group_by(disbursement_month, disbursement_year) %>% 
  summarise(disbursement_amount) %>% 
  arrange(desc(disbursement_amount))

#The highest spending was on month 10th of the year 2018 in the amount of 3390990.85. 

```

Which recipient (recipient_name) was paid the most in the 2020 election cycle?

```{r}

payments %>%
  filter(disbursement_amount == max(disbursement_amount)) 

#I filtered and used the max function to find the largest amount of money that was paid and who 
#was the recipient

#The recipient was the AMERICAN MADE MEDIA CONSULTANTS, LLC. 


```

Look at what this recipient was paid for. What service or good (see the "disbursement_description") were they paid the most for?

```{r}

payments %>% 
  filter(recipient_name=="AMERICAN MADE MEDIA CONSULTANTS, LLC") %>% 
  group_by(disbursement_description,recipient_name,disbursement_amount) %>% 
  arrange(desc(disbursement_amount))

```

What's the most common category overall (see "disbursement_description") the campaign spent money on?

```{r}

payments %>% 
  group_by()
  



```

Who were the 5 biggest recipients of campaign spending on "LEGAL CONSULTING"..?

```{r}

#I filtered for the disbursement description only to show legal consulting
#Then, I selected for it to be grouped by the recipient name and the respective amount being disbursed 
#Then, I selected it to be arranged from descending order of the disbursement to find the highest values 
#Then, I used the function head to select the TOP 5 values 


payments %>% 
  filter(disbursement_description=="LEGAL CONSULTING") %>% 
  group_by(recipient_name, disbursement_amount) %>% 
  arrange(desc(disbursement_amount)) %>% 
  head(5)


# Top 5 biggest recipients of campaign spending on "LEGAL CONSULTING"
  #JONES DAY 	
  #JONES DAY 	
  #JONES DAY 	
  #KASOWITZ, BENSON, TORRESS & FRIEDMAN	
  #JONES DAY

#Those are the Top 5. JONES DAY shows up multiple times because the disbursement dates 
# differ by month and year from when the money is being paid to the recipient


```


Who are the top 5 individuals (vs. organization) were paid by the campaign committee?

```{r}

#I filtered for both the individual and the organization, then I arranged to find the maximum values
#of the disbursement amount by descending order, and used the Head function to find the Top 5 ones. 
#I did two sets to separate individual from organization 

payments %>% 
  filter(entity_type_desc=="INDIVIDUAL") %>% 
  arrange(desc(disbursement_amount)) %>% 
  head(5)

# Top 5 individuals paid by the campaign comittee
  #GIGICOS, GEORGE		
  #CARR, CHRISTOPHER	
  #DOLLMAN, SEAN		
  #GLASSNER, MICHAEL	
  #NEWLAND, DEE EDWARD

payments %>% 
  filter(entity_type_desc=="ORGANIZATION") %>% 
  arrange(desc(disbursement_amount)) %>% 
  head(5)

# Top 5 organizations paid by the campaign comittee
  #AMERICAN MADE MEDIA CONSULTANTS, LLC		
  #ACE SPECIALTIES		
  #AMERICAN EXPRESS	
  #AMERICAN MADE MEDIA CONSULTANTS, LLC		
  #AMERICAN MADE MEDIA CONSULTANTS, LLC



```


For the indvidual who was paid the most, which payment category (disbursement_description) was the top one for him or her? And how much money were they paid in that category?

```{r}

#I filtered for the results only to show the entity description INDIVIDUAL and arranged it in descending 
#order of disbursement. That way, I could find what was the largest amount of money paid to this individual 
#and the category the payment was in. 

payments %>% 
  filter(entity_type_desc=="INDIVIDUAL") %>% 
   arrange(desc(disbursement_amount)) 

#The INDIVIDUAL who was paid the most was Gigicos,George. He was paid 30000.00 for the category of EVENT CONSULTING.  



```
